Determining the existence and direction of causality in the face of measurement error

### 2016-12-28 docs/manuscript.tex
updated

\documentclass[]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Determining the existence and direction of causality in the face of measurement error},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{longtable,booktabs}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\setlength{\droptitle}{-2em}
  \title{Determining the existence and direction of causality in the face of
measurement error}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{}
  \preauthor{}\postauthor{}
  \predate{\centering\large\emph}
  \postdate{\par}
  \date{14 December 2016}

@@

Two sample MR\textsuperscript{19} was performed using the summary
statistics presented in\textsuperscript{18} for each of 458 associations
between DNA methylation and gene expression levels. Because only summary
statistics were available (effect, standard error, effect allele, sample
size, p-values) for the instrumental SNP on the methylation and gene
expression levels, the Steiger test of two independent correlations was
used to infer the direction of causality for each of the associations.
The Wald ratio test was then used to estimate the causal effect size for
the estimated direction for each association.

@@

We performed simulations to verify that this problem does arise using
the CIT method. Figure 1 shows that when there is no measurement error
in the exposure or outcome variables (\(\rho_{x, x_o}=1\)) the CIT is
reliable in identifying the correct causal direction. However, as
measurement error increases in the exposure variable, eventually the CIT
is more likely to infer a robust causal association in the wrong
direction. Also of concern here is that increasing sample size does not
solve the issue, indeed it only strengthens the apparent evidence for
the incorrect inference.

@@

We performed simulations to compare the performance of MR against CIT in
detecting a causal association between simulated variables under
different levels of imprecision simulated in the exposure. Figure 3
shows the true positive rates between the CIT and MR for detecting a
causal association. We observe that the CIT has lower power in all
cases, with performance declining as measurement imprecision increases
in the exposure. The performance of MR in detecting an association is
unrelated to measurement error in. Measurement error in the outcome does
not induce a substantive difference in performance between CIT and MR.

@@

This is a simple method to orient the direction of causality in an MR
analysis when the underlying biology of the SNP is not fully understood.
But it can be shown that in the presence of measurement imprecision,
\(d = \rho_{x, x_o} - \rho_{x,y}\rho_{y,y_o}\) (Appendix 2), where \(d\)
is a metric of the degree to which the Steiger test is liable to provide
the wrong estimate (\emph{i.e.} if \(d>0\) then the Steiger test is
likely to be correct about the causal direction). Thus it is important
to note that under certain conditions when measurement imprecision is
present the Steiger test could make erroneous inference about the causal
direction because the value of \(d\) is not restricted to being greater
than 0. However Figure 2 shows that when there is no measurement error
in \(y\), the Steiger test is unlikely to infer the wrong direction of
causality even if there is measurement error in \(x\). It also shows
that in most cases where \(y\) is measured with error, especially when
the causal effect between \(x\) and \(y\) is not very large, the
condition of \(d > 0\) is satisfied.

We performed simulations to explore the performance of this approach in
comparison to CIT in terms of the rate at which a strong causal
relationship is obtained for the correct direction of causality, and the
rate at which a strong causal relationship is obtained where the
reported direction of causality is incorrect. Figure 4 shows that when
\(d < 0\) the MR analysis is indeed liable to infer the wrong direction
of causality, and that this erroneous result is more likely to occur
with increasing sample size. However, the CIT is in general much more
fallable to reporting a robust causal association for the wrong
direction of causality. When \(d > 0\) is satisfied we observe that in
most cases the MR method has greater power to obtain evidence for
causality than CIT, and always obtains the correct direction of
causality.

@@

We used the Steiger test to test if DNA methylation levels are more
likely to causally influence gene expression levels or vice versa. To do
this we obtained a list of 458 gene expression - DNA methylation
associations as reported in Shakhbazov et al\textsuperscript{18}. These
were filtered to be located on the same chromosome, have robost
correlations after correcting for multiple testing, and to share a SNP
that had a robust cis-acting effect on both the DNA methylation probe
and the gene expression probe. Using the reported summary data we used
the Steiger test of two independent correlations to identify which
direction of causality was most likely for each association. We found
that the causal direction commonly goes in both directions (Table 1) but
methylation levels were more likely to influence expression levels than
vice versa (probability of expression causing methylation = 0.0019317
(95\% CI 0.3455756 - 0.4653544) amongst 272 associations with
\(p_{Steiger} < 0.05\)).

@@

We also observed that for associations where methylation caused gene
expression the causal effect was more likely to be negative than for the
associations where gene expression caused methylation (OR = 0.4615363
(95\% CI 0.259998 - 0.8052745)), suggesting that reducing gene
expression levels at a controlling CpG typically leads to increased gene
expression levels, consistent with expectation\textsuperscript{2}.

@@

\subsection{Tables}\label{tables}

\begin{longtable}[]{@{}lrrrrr@{}}
\toprule
Causal direction & Count & \(p_{Steiger} < 0.05\) & P(+ve effect) &
\(cor(\rho_{MR}, \rho_{P})\) & Mean sensitivity\tabularnewline
\midrule
\endhead
Expression causes Methylation & 246 & 162 & 0.3902439 & 0.5629453 &
0.2031556\tabularnewline
Methylation causes Expression & 212 & 110 & 0.2877358 & 0.4384660 &
0.2365816\tabularnewline
\bottomrule
\end{longtable}

Table 1: Summary of MR analysis on gene expression and DNA methylation
associations. `Count (\(p_{Steiger} < 0.05\))' denotes the number of
associations that have a p-value \textless{} 0.05 from the Steiger test.
P(+ve effect) denotes the proportion of associations that were positive.
\(cor(\rho_{MR}, \rho_{P})\) is the correlation of the MR estimate
obtained in this analysis and the Pearson correlation between reported
in Shakhbazov et al 2016.

@@

Figure 1: The CIT was performed on simulated variables where the
exposure influenced the outcome and the exposure was instrumented by a
SNP. The test statistic from CIT when testing if the exposure caused the
outcome (the true model) is in red, and the test for the outcome causing
the exposure (false model) is in green. Rows of plots represent the
sample sizes used for the simulations. As measurement error increases
(decreasing values on x-axis) the test statistic for the incorrect model
gets stronger and the test statistic for the correct model gets weaker.

@@

\includegraphics{manuscript_files/figure-latex/d_relationship_figure-1.pdf}

Figure 2: Plots depicting the parameter space in which the function
\(d = cor(x, x_o) - cor(x,y)cor(y, y_o)\) is negative. When \(d\) is
negative the Steiger test is liable to infer the wrong direction of
causality. \(d\) is plotted on the y axis, with measurement precision of
\(x\) (\(cor(x,x_o)\)) on the x-axis. Higher measurement precision of
\(y\) (\(cor(y,y_o)\)) is depicted by lighter coloured lines, and
columns of boxes are for different true values of \(cor(x,y)\). Shaded
regions show the parameter space where \(d\) is negative. The graph
shows that for the majority of the parameter space of the function,
\(d\) is positive, especially where causal relationships are relatively
weak.

@@

Figure 3: Outcomes were simulated to be causally influenced by exposures
with varying degrees of measurement imprecision applied to the exposure
variable (x axis). True positive rates (y axis) for MR and CIT were
compared with for varying levels of measurement imprecision in the
outcome variable (rows of boxes) and sample sizes (columns of boxes).

@@

Figure 4: Outcomes were simulated to be causally influenced by
exposures, with varying degrees of measurement error applied to both.
CIT and MR were used to infer evidence for causality between the
exposure and outcome, and to infer the direction of causality. The value
of \(d = \rho_{x, x_o} - \rho_{x,y}\rho_{y,y_o}\), such that when \(d\)
is negative we expect the Steiger test to be more likely to be wrong
about the direction of causality. Rows of graphs represent the sample
size used in the simulations. For the CIT method, outcome 1 denoted
evidence for causality with correct model, outcomes 2 or 3 denoted
evidence for causality with incorrect model, and outcome 4 denoted no
evidence for causality.

@@

where \(\hat{x} = \hat{\alpha}_1 + \hat{\beta}_1 g\). Writing in scale
free terms, \(\rho_{g, x}\) denotes the correlation between \(g\) and
the exposure variable \(x\), and it is expected that
\(\rho_{g, x} > \rho_{g, y}\) because
\(\rho_{g, y} = \rho_{g, x}\rho_{x, y}\), where \(\rho_{x, y}\) is the
causal association between \(x\) and \(y\) (which is likely to be less
than 1). In the presence of measurement error in \(x\) and \(y\),
however, Steiger test will instead be assessing the inequality
\(\rho_{g, x_o} > \rho_{g, y_o}\), which can be simplified:

\[
\begin{aligned}
\rho_{g, x_o} & > \rho_{g, y_o} \\
\rho_{g, x} \rho_{x, x_o} & > \rho_{g,y}\rho_{y,y_o}\\
\rho_{g, x} \rho_{x, x_o} & > \rho_{g,x}\rho_{x,y}\rho_{y,y_o}\\
\rho_{x, x_o} & > \rho_{x,y}\rho_{y,y_o}
\end{aligned}
\]

\subsection{Appendix 3}\label{appendix-3}

The Steiger test is susceptible to inferring the wrong direction of
causality under some measurement error parameters. For an observed
association between the instrument \(g\) and the measured exposure
\(x_o\), \(\rho_{g,x_o}\), the association between \(g\) and the true
exposure (without measurement error) \(x\),
\(\rho_{g,x} = \frac{\rho_{x,x_o}}{\rho_{g,x_o}}\) where the bounds
\(\rho_{g,x_o} \leq \rho_{x,x_o} \leq 1\) are the theoretical limits of
the correlation between true and measured \(x\). The same logic applies
to the association between the instrument and the outcome variable.

For the values \(\rho_{g,x_o} = 0.3\) and \(\rho_{g,y_o} = 0.05\), the
possible true \(\rho_{g,x}\) and \(\rho_{g,y}\) enumerated for all
possible values of measurement error can be visualised as so:

\includegraphics{manuscript_files/figure-latex/unnamed-chunk-1-1.pdf}

Here, where \(z = \rho_{g,y} - \rho_{g,x} < 0\) the measurement error
range in \(x_o\) and \(y_o\) will not change the qualitative inference
of the direction of causality. To evaluate the sensitivity, \(S\) of the
Steiger test to measurement error, the objective is to calculate the
proportion of the total absolute volume of this function that is above
the \(z = 0\) plane,

@@

\(V_{z}\), the total volume of the function, can be obtained
analytically by solving the double integral:

@@

\(S\) can take values from 0 to 0.5, where low values suggest low
sensitivity of the Steiger test to measurement error. For the values
used to plot the function above, \(\rho_{g,x_o} = 0.3\) and
\(\rho_{g,y_o} = 0.05\), we obtain \(S = 0.032\). Hence, the majority of
the parameter space of measurement error will not change the inferred
causal direction of \(x \rightarrow y\) for these values.

>>>>

\documentclass[]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Determining the existence and direction of causality in the face of measurement error},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\setlength{\droptitle}{-2em}
  \title{Determining the existence and direction of causality in the face of
measurement error}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{}
  \preauthor{}\postauthor{}
  \predate{\centering\large\emph}
  \postdate{\par}
  \date{28 December 2016}

@@

Two sample MR\textsuperscript{19} was performed using the summary
statistics presented in\textsuperscript{18} for each of 4122
associations between DNA methylation and gene expression levels. Because
only summary statistics were available (effect, standard error, effect
allele, sample size, p-values) for the instrumental SNP on the
methylation and gene expression levels, the Steiger test of two
independent correlations was used to infer the direction of causality
for each of the associations. The Wald ratio test was then used to
estimate the causal effect size for the estimated direction for each
association.

@@

We performed simulations to verify that this problem does arise using
the CIT method. Figure 2 shows that when there is no measurement error
in the exposure or outcome variables (\(\rho_{x, x_o}=1\)) the CIT is
reliable in identifying the correct causal direction. However, as
measurement error increases in the exposure variable, eventually the CIT
is more likely to infer a robust causal association in the wrong
direction. Also of concern here is that increasing sample size does not
solve the issue, indeed it only strengthens the apparent evidence for
the incorrect inference.

@@

We performed simulations to compare the performance of MR against CIT in
detecting a causal association between simulated variables under
different levels of imprecision simulated in the exposure. Figure 4
shows the true positive rates between the CIT and MR for detecting a
causal association. We observe that the CIT has lower power in all
cases, with performance declining as measurement imprecision increases
in the exposure. The performance of MR in detecting an association is
unrelated to measurement error in. Measurement error in the outcome does
not induce a substantive difference in performance between CIT and MR.

@@

This is a simple method to orient the direction of causality in an MR
analysis when the underlying biology of the SNP is not fully understood.
But it can be shown that in the presence of measurement imprecision,
\(d = \rho_{x, x_o} - \rho_{x,y}\rho_{y,y_o}\) (Appendix 2), where \(d\)
is a metric of the degree to which the Steiger test is liable to provide
the wrong estimate (\emph{i.e.} if \(d>0\) then the Steiger test is
likely to be correct about the causal direction). Thus it is important
to note that under certain conditions when measurement imprecision is
present the Steiger test could make erroneous inference about the causal
direction because the value of \(d\) is not restricted to being greater
than 0. However Figure 7 shows that when there is no measurement error
in \(y\), the Steiger test is unlikely to infer the wrong direction of
causality even if there is measurement error in \(x\). It also shows
that in most cases where \(y\) is measured with error, especially when
the causal effect between \(x\) and \(y\) is not very large, the
condition of \(d > 0\) is satisfied.

We performed simulations to explore the performance of this approach in
comparison to CIT in terms of the rate at which a strong causal
relationship is obtained for the correct direction of causality, and the
rate at which a strong causal relationship is obtained where the
reported direction of causality is incorrect. Figure 5 shows that when
\(d < 0\) the MR analysis is indeed liable to infer the wrong direction
of causality, and that this erroneous result is more likely to occur
with increasing sample size. However, the CIT is in general much more
fallable to reporting a robust causal association for the wrong
direction of causality. When \(d > 0\) is satisfied we observe that in
most cases the MR method has greater power to obtain evidence for
causality than CIT, and always obtains the correct direction of
causality.

@@

We used the Steiger test to infer the direction of causality between DNA
methylation levels and gene expression levels. To do this we obtained a
list of 458 gene expression - DNA methylation associations as reported
in Shakhbazov et al\textsuperscript{18}. These were filtered to be
located on the same chromosome, have robost correlations after
correcting for multiple testing, and to share a SNP that had a robust
cis-acting effect on both the DNA methylation probe and the gene
expression probe. Using the reported summary data we used the Steiger
test of two independent correlations to identify which direction of
causality was most likely for each association. We found that the causal
direction commonly goes in both directions (Table 1) but methylation
levels were more likely to influence expression levels than vice versa
(probability of expression causing methylation = 0.0163873 (95\% CI
0.3738727 - 0.4874269) amongst NA associations with
\(p_{Steiger} < 0.05\)).

@@

We also observed that for associations where methylation caused gene
expression the causal effect was more likely to be negative than for the
associations where gene expression caused methylation (OR = 0.5326663
(95\% CI 0.3172203 - 0.8851868)), suggesting that reducing gene
expression levels at a controlling CpG typically leads to increased gene
expression levels, consistent with expectation\textsuperscript{2}.

@@

\includegraphics{../images/dag-01.png}

Figure 1: Gene expression levels (blue blocks) and DNA methylation
levels (green triangles) may be correlated but the causal direction is
unknown. If a SNP (yellow circle) is associated with both DNA
methylation and gene expression levels then it can be used as an
instrument, but there are two competing models for these variables. a)
Methylation causes gene expression. The left figure shows that the SNP
influences methylation levels that in turn influence gene expression
levels. The right figure shows the directed acyclic graph that
represents this model. Faded symbols represent the measured values
whereas solid symbols represent the true values. b) The same as in A,
except the causal direction is from gene expression to DNA methylation.

@@

Figure 2: The CIT was performed on simulated variables where the
exposure influenced the outcome and the exposure was instrumented by a
SNP. The test statistic from CIT when testing if the exposure caused the
outcome (the true model) is in red, and the test for the outcome causing
the exposure (false model) is in green. Rows of plots represent the
sample sizes used for the simulations. As measurement error increases
(decreasing values on x-axis) the test statistic for the incorrect model
gets stronger and the test statistic for the correct model gets weaker.

@@

\includegraphics{manuscript_files/figure-latex/steiger_sensitivity_plot-1.pdf}

Figure 3: a) We can predict the values the Steiger test would take
(z-axis) for different potential values of measurement error (x and y
axes), drawn here as the blue surface. When \(\rho_{g,y} > \rho_{g,x}\),
as denoted by range of values where the blue surface is above the red
plane, those values of measurement error lead to our observed Steiger
test inferring the wrong causal direction. Where the blue surface lies
below the red plane, these measurement error values support the inferred
causal direction of X to Y. A measure of sensitivity, therefore, is the
volume of the positive space bound by the blue surface as a proportion
of the total space bound by the blue surface. Sensitivity of 0 denotes
no measurement error values that could lead to inferrence of incorrect
direction, whereas the maximum value of 0.5 denotes measurement error
equal chance of either direction across all measurement error ranges. In
this case, where \(\rho_{g,x}^2 = 0.01\) and \(\rho_{x,y}^2 = 0.1\), the
sensitivity is 0.18. b) The influence of \(\rho_{x,y}\) (x-axis) on the
Steiger sensitivity values (y-axis) for different values of
\(\rho_{g,x}^2\). Inferring the direction of causality is more
susceptible to measurement error when there are stronger causal
associations between X and Y.

@@

Figure 4: Outcomes were simulated to be causally influenced by exposures
with varying degrees of measurement imprecision applied to the exposure
variable (x axis). True positive rates (y axis) for MR and CIT were
compared for varying levels of measurement imprecision in the outcome
variable (rows of boxes) and sample sizes (columns of boxes).

@@

Figure 5: Outcomes were simulated to be causally influenced by
exposures, with varying degrees of measurement error applied to both.
CIT and MR were used to infer evidence for causality between the
exposure and outcome, and to infer the direction of causality. The value
of \(d = \rho_{x, x_o} - \rho_{x,y}\rho_{y,y_o}\), such that when \(d\)
is negative we expect the Steiger test to be more likely to be wrong
about the direction of causality. Rows of graphs represent the sample
size used in the simulations. For the CIT method, outcome 1 denoted
evidence for causality with correct model, outcomes 2 or 3 denoted
evidence for causality with incorrect model, and outcome 4 denoted no
evidence for causality.

\includegraphics{manuscript_files/figure-latex/shakhplot-1.pdf}

Figure 6: Using 458 putative associations between DNA methylation and
gene expression we used the Steiger test to infer the direction of
causality between them. The rightmost bar shows the proportion of
associations for each of the two possible causal directions (colour key)
assuming no measurement error in either gene expression or DNA
methylation levels. The proportions change when we assume different
levels of measurement error in gene expression levels (x-axis) or DNA
methylation levels (columns of boxes). If there is systematically higher
measurement error in one platform it will appear to be less likely to be
the causal factor.

@@

where \(\hat{x} = \hat{\alpha}_1 + \hat{\beta}_1 g\). Writing in scale
free terms, \(\rho_{g, x}\) denotes the correlation between \(g\) and
the exposure variable \(x\), and it is expected that
\(\rho_{g, x} > \rho_{g, y}\) because
\(\rho_{g, y} = \rho_{g, x}\rho_{x, y}\), where \(\rho_{x, y}\) is the
causal association between \(x\) and \(y\) (which is likely to be less
than 1).

In the presence of measurement error in \(x\) and \(y\), however, the
Steiger test will instead be assessing the inequality
\(\rho_{g, x_o} > \rho_{g, y_o}\). In order to assess how sensitive the
Steiger test is to measurement error, we can evaluate the range of
potential values of measurement error in \(x\) and \(y\) over which the
empirical result from the Steiger test would return the wrong value.

For different values of \(\rho_{x,x_o}\),
\(\rho_{g,x} = \frac{\rho_{g, x_o}}{\rho_{x,x_o}}\) and
\(\rho_{g,x_o} \leq \rho_{x,x_o} \leq 1\). For different values of
\(\rho_{y,y_o}\), \(\rho_{g,y} = \frac{\rho_{g, y_o}}{\rho_{y,y_o}}\)
and \(\rho_{g,y_o} \leq \rho_{y,x_o} \leq 1\).

Where \(z = \rho_{g,y} - \rho_{g,x} < 0\) the measurement error range in
\(x_o\) and \(y_o\) will not change the qualitative inference of the
direction of causality. To evaluate the sensitivity, \(S\), of the
Steiger test to measurement error, the objective is to calculate the
proportion of the total absolute volume of this function that is above
the \(z = 0\) plane,

@@

\(V_{z}\), the total volume of the function (Figure 3), can be obtained
analytically by solving:

@@

The sensitivity, \(S\), can take values from 0 to 0.5, where low values
suggest there are
